---
output: github_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# personMatchR


Helper package for matching individuals across two data-sets. This R package has been developed by 
the NHS Business Services Authority Data Science team.

The matching functions will identify whether a record containing person information can be matched 
to any records in a second data-set, either based on identical person information or a close match 
based on similar information.

The matching process focuses on comparing individuals based on four key pieces of information: 

* Forename
* Surname
* Date of Birth
* Postcode

The personMatchR package has different functions available to handle matching, whether the input 
data is held within data frames or via a connection to database tables:

* calc_match_patients
  + suitable for data-frames containing up to one million records
* calc_match_patients_db
  + currently only set up and tested for Oracle database infrastructure used by NHSBSA
  + suitable for large volume data-sets
  + data formatting will need to be handling prior to matching, with functions in available in 
  package to support this



## Installation

You can install the development version of personMatchR from the NHSBSA Data Analytics [GitHub](https://github.com/nhsbsa-data-analytics/) with:

``` r
# install.packages("devtools")
devtools::install_github("nhsbsa-data-analytics/personMatchR")
```

## Documentation

In additional to function help files, some additional documentation has been included to provide 
detail information about the package functions and some examples of the package in use:

* [Package Usage Blog](documentation/personMatchR Usage Blog.pdf)
  + Basic overview of package including example test case
* [Quality Review](documentation/personMatchR - Quality Review.pdf)
  + Summary of test case used to check quality of matching process
  + From 37 million records checked, the correct person was identified in 99.9% of cases
* [Example Use Case](documentation/personMatchR - Use Case.pdf)
  + Summary of real world use case using package to match service users to an additional data-set 
  to supplement data
  + Of 31.5 thousand service users, 99% could be aligned to the secondary data-set

## Example

The following basic example shows how to match between two data-frames that are available within 
the documentation folder for this package.

```{r example_data}
df_A <- readRDS("documentation/TEST_DF_A.rds")
head(df_A)
df_B <- readRDS("documentation/TEST_DF_B.rds")
head(df_B)
```

With such a small set of data it is easy to manual compare these two data-sets, where it is clear 
that records are similar between both data-sets:

* Record 1 in both data-sets only differs by an apostrophe in the surname field
* Record 2 in both data-sets has a different version of the forename
* Record 3 in data-set A does not appear anywhere in data-set B
* Record 4 in both data-sets only differs by the forename and surname being swapped

Matching these data-sets we would hope that records 1, 2 and 4 are matched.

We can pass the data-sets to the calc_match_patients function and review the output. For this 
example we will set parameters to only return the key fields from the matching, format the data 
prior to matching and include records without a match in the output:

```{r example, results = FALSE}
library(personMatchR)
library(dplyr)
df_output <- personMatchR::calc_match_patients(
  df_one = df_A, # first data-set
  id_one = ID, # unique id field from first data-set
  forename_one = FORENAME, # forename field from first data-set
  surname_one = SURNAME, # surname field from first data-set
  dob_one = DOB, # date of birth field from first data-set
  postcode_one = POSTCODE, # postcode field from first data-set
  df_two = df_B, # second data-set
  id_two = ID, # unique id field from second data-set
  forename_two = FORENAME, # forename field from second data-set
  surname_two = SURNAME, # surname field from second data-set
  dob_two = DOB, # date of birth field from second data-set
  postcode_two = POSTCODE, # postcode field from second data-set
  output_type = "key", # only return the key match results
  format_data = TRUE, # format input data-sets prior to matching
  inc_no_match = TRUE # return records from first data-set without matches
)
```

```{r example_output, echo=FALSE}
head(df_output)
```


### Understanding match output: MATCH_COUNT & MATCH_SCORE

These fields in the output provide context for the match results:
*The match results show that the results are as we would have expected based on the input data, with the 
output including some fields to provide additional context:

* MATCH_COUNT
  + Shows the number of matches found for each record (each match will be included in the output)
  + Where this is greater than one, some additional handling may be required to review
* MATCH_SCORE
  + Provides a general weighted score for the match which may help compare instances where 
multiple potential matches are identified. 
  + please note that this is for guide purposes only and a higher score will not always mean that 
  the match is more likely to be correct. Weightings for each part of the matching can be adjusted 
  using parameters in the function call.


### Match function parameter: output_type

This parameter will determine the number of fields from each data-set returned in the output:

* key
  + this option will only return the ID fields from each data-set, plus the match outcome and match score.
  + this output could then be used to join to original data-sets as required
* match
  + this option will return the "key" fields plus the personal information fields used in the matching
  + this output would allow the matches to easily be reviewed
* all
  + will return the key fields, plus all other fields from both data-sets where a match was found 


### Match function parameter: format_data (not used for database match functions)

This parameter will determine whether or not the data is formatted as it is passed to the matching 
function. Formatting the data can help ensure both data-sets are consistently formatted, accounting 
for things like case, removal of special characters, date of birth and postcode patterns. 
As formatted data is likely to have better matching outcomes it is strongly advised to apply this 
option.

* TRUE
  + (RECOMMENDED) apply formatting to input data
* FALSE
  + do not apply formatting to input data
  + this is likely to limit match results
  + formatting functions (format_name, format_dob, format_dob) could be applied individually 
  outside of the matching function


### Match function parameter: inc_no_match

This parameter will determine whether or not the output results will include details of records 
where no match could be found:

* TRUE
  + all records from the initial data-set will be included in the output, even if no match was found
* FALSE
  + the output will only include records where a match could be identified
