---
output: github_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# personMatchR


Helper package for matching individuals across two datasets. This R package has been developed by 
the NHS Business Services Authority Data Science team.

The package is a series of functions that process and then match person information fields across 
two datasets, either in data frames or database tables.

The matching functions will identify whether a record containing person information can be matched 
to any records in a second dataset, either based on identical person information or a close match 
based on similar information.

The matching process focuses on comparing individuals based on four key pieces of information: 

* Forename
* Surname
* Date of Birth
* Postcode

The quality of matching accuracy will be heavily influenced by the formatting of the input data, and 
the are processing functions available within the package to support this ([see data preparation](#data-preparation)).

The personMatchR package has different functions available to handle matching, whether the input 
data is held within data frames or via a connection to database tables:

* calc_match_person
  + suitable for data-frames containing up to one million records
* calc_match_person_db
  + currently only set up and tested for Oracle database infrastructure used by NHSBSA
  + suitable for large volume datasets
  + data formatting will need to be handled prior to matching, with functions available in the package to support this



## Installation

You can install the development version of personMatchR from the NHSBSA Data Analytics [GitHub](https://github.com/nhsbsa-data-analytics/) with:

``` r
# install.packages("devtools")
devtools::install_github("nhsbsa-data-analytics/personMatchR")
```

## Documentation

In addition to function help files, some additional documentation has been included to provide 
detailed information about the package functions and some examples of the package in use:

* [Package Usage Blog](documentation/personMatchR Usage Blog.pdf)
  + Basic overview of package, including example test case


## Requirements

The datasets being matched each require a forename, surname, DOB and postcode field to be present. 
In addition, each dataset also requires a unique identification field to be present. Users will have 
to generate such a field prior to using the matching function if one is not already present.


## Example

The following basic example shows how to match between two data-frames that are available within 
the [documentation folder](documentation/) for this package.

```{r example_data}
df_A <- readRDS("documentation/TEST_DF_A.rds")
head(df_A)
df_B <- readRDS("documentation/TEST_DF_B.rds")
head(df_B)
```

With such a small set of data it is easy to manually compare these two datasets, where it is clear 
that records are similar between both datasets:

* Record 1 in both datasets only differs by an apostrophe in the surname field
* Record 2 in both datasets only has a different version of the forename
* Record 3 in dataset A does not appear anywhere in dataset B
* Record 4 in both datasets only differs by the forename and surname being swapped

When matching these datasets we would hope that records 1, 2 and 4 are matched.

We can pass the datasets to the calc_match_person function and review the output. For this 
example we will set parameters to only return the key fields from the matching, format the data 
prior to matching and include records without a match in the output:

```{r example, results = FALSE}
library(personMatchR)
library(dplyr)
df_output <- personMatchR::calc_match_person(
  df_one = df_A, # first dataset
  id_one = ID, # unique id field from first dataset
  forename_one = FORENAME, # forename field from first dataset
  surname_one = SURNAME, # surname field from first dataset
  dob_one = DOB, # date of birth field from first dataset
  postcode_one = POSTCODE, # postcode field from first dataset
  df_two = df_B, # second dataset
  id_two = ID, # unique id field from second dataset
  forename_two = FORENAME, # forename field from second dataset
  surname_two = SURNAME, # surname field from second dataset
  dob_two = DOB, # date of birth field from second dataset
  postcode_two = POSTCODE, # postcode field from second dataset
  output_type = "key", # only return the key match results
  format_data = TRUE, # format input datasets prior to matching
  inc_no_match = TRUE # return records from first dataset without matches
)
```

```{r example_output, echo=FALSE}
head(df_output)
```

The match results show exact matches for records 1 and 4 as the only differences were special 
characters and transposition of names. For record 2 the results show a confident match, as although 
not identical the names were similar enough to pass the confidence thresholds. As expected, record 
3 does not produce any matches.

### Understanding match output: MATCH_COUNT & MATCH_SCORE

These fields in the output provide context for the match results:

* MATCH_COUNT
  + Shows the number of matches found for each record (each match will be included in the output).
  + Where this is greater than one, some additional handling may be required to review.
* MATCH_SCORE
  + Provides a general weighted score for the match which may help compare instances where 
multiple potential matches are identified. 
  + Please note that this is for guide purposes only, and a higher score will not always mean that 
  the match is more likely to be correct. Weightings for each part of the matching can be adjusted 
  using parameters in the function call.


## Data preparation

In the example above the input data was passed through some formatting functions as part of the main 
matching package function call (format_data = TRUE). This option is only available when matching 
across data frames using the calc_match_person() function. However, the formatting functions could 
be called individually prior to calling the matching function.

There are three matching functions available, with different versions available for the database 
matching function:

* format_name()       / format_name_db()
* format_date()       / format_date_db()
* format_postcode()   / format_postcode_db()

The following code shows how these functions could be used to format the data prior to matching:
```{r formatting example}
df_A = df_A %>% 
  format_date(date = DOB) %>% 
  format_name(name = FORENAME) %>% 
  format_name(name = SURNAME) %>% 
  format_postcode(id = ID, postcode = POSTCODE)
head(df_A)
```

The calc_match_person_db() function **does not** offer the option to format the data prior to matching, 
with users required to carry out the processing as above beforehand. 

It is **strongly** encouraged that users create new database tables after processing these fields. 
They can then use these freshly created tables as the matching function input, which will typically 
significantly improve runtime performance.


## Match function parameter: output_type

This parameter will determine the number of fields from each dataset returned in the output:

* key
  + this option will only return the ID fields from each dataset, plus the match outcome and match score
  + this output could then be used to join to original datasets as required
* match
  + this option will return the "key" fields plus the personal information fields used in the matching
  + this output would allow the matches to easily be reviewed
* all
  + will return the key fields, plus all other fields from both datasets where a match was found 


## Match function parameter: format_data (not used for database match functions)

This parameter will determine whether or not the data is formatted as it is passed to the matching 
function. Formatting the data can help ensure both datasets are consistently formatted, accounting 
for things like case, removal of special characters, date of birth and postcode patterns. 
As formatted data is likely to have better matching outcomes it is strongly advised to apply this 
option.

* TRUE
  + (RECOMMENDED) apply formatting to input data
* FALSE
  + do not apply formatting to input data
  + this is likely to limit match results
  + formatting functions (format_name, format_dob, format_dob) could be applied individually 
  outside of the matching function, prior to passing data to main matching function


## Match function parameter: inc_no_match

This parameter will determine whether or not the output results will include details of records 
where no match could be found:

* TRUE
  + all records from the initial dataset will be included in the output, even if no match was found
* FALSE
  + the output will only include records where a match could be identified
